---
title: "Applicability domain methods for continuous data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{continuous-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 100)
library(ggplot2)
theme_set(theme_bw())
```

## Introduction

```{r}
library(applicable)
```

`applicable` provides the following methods to analyze the applicability domain of your model:

* Principal component analysis
* Hat values statistics

## Example

We will use the Ames IA housing data for our example.

```{r ames_data, message=FALSE}
library(AmesHousing)
ames <- make_ames()
```

There are `r format(nrow(ames), big.mark = ",")` properties in the data.

The Sale Price was recorded along with `r ncol(ames)` predictors, including:

* Location (e.g. neighborhood) and lot information.
* House components (garage, fireplace, pool, porch, etc.).
* General assessments such as overall quality and condition.
* Number of bedrooms, baths, and so on.

More details can be found in [De Cock (2011, Journal of Statistics Education)](http://ww2.amstat.org/publications/jse/v19n3/decock.pdf).

The raw data are at [`http://bit.ly/2whgsQM`](http://bit.ly/2whgsQM) but we 
will use a processed version found in the 
[`AmesHousing`](https://github.com/topepo/AmesHousing) package. `applicable` 
also contains an update for these data for three new properties (although fewer 
fields were collected on these). 

To pre-process the training set, we will use the _recipes_ package. We first
tell the recipes that there is an additional value for the neighborhood in these
data, then direct it to create dummy variables for all categorical predictors. In
cases where there are no levels observed for a factor, we eliminate predictors
with a single unique value, then estimate a transformation that will make the
predictor distributions more symmetric. After these, the data are centered and
scaled. These same transformations will be applied to the new data points using
the statistics estimated from the training set.

```{r prep_data, message=FALSE}
library(recipes)
library(dplyr)

ames_cols <- names(ames_new)

training_data <- 
  ames %>% 
  # For consistency, only analyze the data on new properties
  dplyr::select(one_of(ames_cols)) %>% 
  mutate(
    # There is a new neighborhood in ames_new
    Neighborhood = as.character(Neighborhood),
    Neighborhood = factor(Neighborhood, levels = levels(ames_new$Neighborhood))
  )


training_recipe <-
  recipe( ~ ., data = training_data) %>%
  step_dummy(all_nominal()) %>% 
  # Remove variables that have the same value for every data point.
  step_zv(all_predictors()) %>% 
  # Transform variables to be distributed as Gaussian-like as possible.
  step_YeoJohnson(all_numeric()) %>%
  # Normalize numeric data to have a mean of zero and
  # standard deviation of one.
  step_normalize(all_numeric())
```


### Principal Component Analysis

The following functions in `applicable` are used for principal component
analysis:

* `apd_pca`: computes the principal components that account for up
to either 95% or the provided `threshold` of variability. It also computes the
percentiles of the principal components and the mean of each principal
component.
* `autoplot`: plots the distribution function for pcas. You can also provide an
optional set of `dplyr` selectors, such as `dplyr::matches()` or
`dplyr::starts_with()`, for selecting which variables should be shown in the
plot.
* `score`: calculates the principal components of the new data and their
percentiles as compared to the training data. The number of principal
components computed depends on the `threshold` given at fit time. It also
computes the multivariate distance between each principal component and its
mean.

Let us apply `apd_pca` modeling function to our data:

```{r}
ames_pca <- apd_pca(training_recipe, training_data)
ames_pca
```

Since no `threshold` was provided, the function computed the number of
principal components that accounted for at most 95% of the total variance.

For illustration, setting `threshold = 0.25` or 25%, we now need only 10 principal components:

```{r}
ames_pca <- apd_pca(training_recipe, training_data, threshold = 0.25)
ames_pca
```

Plotting the distribution function for the PCA scores is also helpful:

```{r autoplot}
library(ggplot2)
autoplot(ames_pca)
```

You can use regular expressions to plot a smaller subset of the pca statistics:

```{r, echo = FALSE, results = "hold"}
autoplot(ames_pca, matches("PC0[1-5]"))
autoplot(ames_pca, distance) + scale_x_log10()
```

The `score` function compares the training data to new samples. Let's go back to 
the case where we capture 95% of the variation in the predictors and score the 
new samples. Since we used the recipe interface, we can give the score function
the original data:

```{r new_sample}
ames_pca <- apd_pca(training_recipe, training_data)
pca_score <- score(ames_pca, ames_new)
pca_score %>% select(matches("PC00[1-3]"), contains("distance"))
```

Notice how the samples, displayed in red, are fairly disimilar to the training
set in the first component:

```{r, echo = FALSE}
training_scores <- score(ames_pca, training_data)
ggplot(training_scores, aes(x = PC001)) + 
  geom_histogram(col = "white", binwidth = .5) + 
  geom_vline(xintercept = pca_score$PC001, col = "red")
```

What is driving the first component? We can look at which predictors have the
largest values in the rotation matrix (i.e. the values that define the linear 
combinations in the PC scores). The top five are:

```{r}
# `ames_pca$pcs` is the output of `prcomp()`
comp_one <- ames_pca$pcs$rotation[, 1]
comp_one[order(abs(comp_one), decreasing = TRUE)] %>% head(5)
```

These three houses are extreme in the most influential variable (year built) since
they were new homes. The also tend to have fairly large garages:

```{r, echo = FALSE}
ggplot(training_data, aes(x = Garage_Area )) + 
  geom_histogram(col = "white", binwidth = 50) + 
  geom_vline(xintercept = ames_new$Garage_Area, col = "red")
```

This may be what is driving the first component. 

However, the overall distance values are relatively small, which indicates that,
overall, these new houses are not outside the mainstream of the data. 



### Hat Values

The [Hat or leverage values](https://en.wikipedia.org/wiki/Leverage_(statistics)) are based on the numerics of linear regression. The measure the distance of a data point to the center of the training set distribution. For example, if the numeric training set matrix was $X_{n \times p}$, the _hat matrix_ for the training set would be computed using 

$$H = X'(X'X)^{-1}X$$

The corresponding hat values for the training would be the diagonals of $H$. These values can be computed using `stats::hatvalues(lm_model)` but only for an `lm` model object. Also, it cannot compute the values for new samples. 

Suppose that we had a new, unknown sample (as a $p \times 1$ data vector $u$). The hat value for this sample would be 

$$h = u'(X'X)^{-1}u$$.

The following functions in `applicable` are used to compute the hat values of your model:

* `apd_hat_values`:
* `score`:

Two caveats for using the hat values:

 1. The numerical methods are less tolerant than PCA. For example, extremely correlated predictors will degrade the ability of the hat values to be effectively used. Also, since an inverse is used, there cannot be an linear dependencies within $X$. To resolve this the former example, the recipe step `recipes::step_corr()` can be used to reduce correlation. For the latter issue, `recipes::step_lincomp()` will idnetify and remove linear dependencies in the data (as shown below). 
 
 1. When using a linear or logistic model, the model adds an intercept columns of ones to $X$. For equivalent computations, you should add a vector or ones to the data or use `recipes::step_intercept()`. 

Let us apply `apd_hat_values` modeling function to our data (while ensuring that there are no linear dependencies):

```{r}
non_singular_recipe <- 
  training_recipe %>% 
  step_lincomb(all_predictors())

# Recipe interface
ames_hat <- apd_hat_values(non_singular_recipe, training_data)
```

